---
title: "Data Profiling"
subtitle: "Preliminary QA"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    theme: simplex
    toc: true
---

# Introduction

The steps taken in this notebook are:

- Read data into R (as a tibble)
- Make column names R-friendly
- Check variable types
- Check missing values
- Range calculations
- Count calculations

```{r setup}
#| include: false

source("utils.R")
```

# Read data

For this example we will use the `maven_market.xlsx` dataset which contains data about product inventory for the fictional company Maven Market.

```{r read-data}

# Read data into a tibble
inventory_path <- here("02_data/raw/maven_market.xlsx")

inventory_raw <- 
  read_excel(
    inventory_path,
    sheet = "Preliminary QA",
    range = "B3:H103",
    trim_ws = FALSE
  )

# View the head of the tibble
head(inventory_raw)
```

**Note on the code above:** normally I would leave the `trim_ws` argument in the `read_excel()` function set to its default value of `TRUE`. This will ensure any whitespace at the start or end of a string is removed. But for the purposes of showing the full data QA steps here, I have opted not to use it.

# Make column names R-friendly

This is a simple step using the `janitor` package which converts the column headers to a more code-friendly format. This helps when writing the rest of the code. Note, when I first read in the data, I stored it in the variable `inventory_raw`. I always keep a copy of the raw, unprocessed data in a variable with the `_raw` suffix so that I can compare the results of any cleaning actions against the original data. Now that I am starting to clean the data, I store it in a new variable called `inventory_clean`.

```{r clean-names}

# Clean the column names
inventory_clean <- inventory_raw |> 
  clean_names()
```

## Glimpse

At this point, it makes sense to use the `glimpse()` function to see the new column headings and to get some more information about the data.

```{r glimpse-data}
glimpse(inventory_clean)
```

# Check variable types

Looking at the output from `glimpse()` above, I notice the following issues:

- The `stock_date` column is incorrectly stored as a character variable. This should be stored as a date.
- The `product_id_sku` column is incorrectly stored as numeric. This should be a character variable.

I'm curious about the `stock_date` column. The values that we can see from `glimpse()` all look like numbers but are stored as text. If the entire column were numbers, then R would have read this in as numeric. Let's take a closer look at all of the distinct values in this column.

```{r stock-date-distinct}
#| paged-print: true

# View the distinct values of stock_date and how many times they occur

inventory_clean |> 
  count(stock_date)
```

OK, so there are some values where the date has been entered as a text string. To fix this and convert all the values to dates, I'll need to use the `lubridate` and `janitor` packages and fix this in two steps. The key is to first try parsing the dates in the `dd-mmm-yy` format using `lubridate::dmy()`. If that fails (which it will for the numeric strings), I then process those failed cases as Excel numeric dates using `janitor::excel_numeric_to_date()`.

```{r fix-variable-types}

# Fix the incorrect variable types
inventory_clean <- 
  inventory_clean |> 
  mutate(
    stock_date = if_else(
      is.na(dmy(stock_date)),
      excel_numeric_to_date(as.numeric(stock_date)),
      dmy(stock_date)
    ),
    product_id_sku = as.character(product_id_sku)
  )
```
Let's check the results.

## Stock date

```{r check-results-stock-date}
summary(inventory_clean$stock_date)
```

I can see this is now being treated correctly as a date variable and there are no empty values (if there were any empty values, the `summary()` function would include a column for the number of NAs.

## Product ID

```{r check-results-product-id}
summary(inventory_clean$product_id_sku)
```

This column is now a character column with 100 values.

# Check for missing values

The code chunk below returns the number of missing values in each column of the dataframe.

```{r count-missing-values}

# Return a count per column of missing values
colSums(is.na(inventory_clean))
```

There are no missing values in any of the columns of the dataframe.

# Range calculations

The `skimr` package contains a nice function to summarise each column.

```{r skimr-view-of-data}
# Use skimr to profile the data
skim(inventory_clean)
```

Looking at the ranges in this summary I notice the following:

- Unit prices range from 0 to 375. The histogram shows that the 375 is a real outlier, and knowing what this business sells it's unlikely that there would be a true unit price of 375. It's more likely that this should be 3.75.
- There shouldn't be any products with a 0 unit price either. These will need to either be imputed or removed from the dataset. Since I don't know what the correct value is, I will remove these rows from the dataframe.

```{r fix-incorrect-unit-prices}

# Correct the unit prices where necessary
inventory_clean <- 
  inventory_clean |> 
  mutate(
    unit_price = case_when(
      unit_price == 0 ~ NA,
      product_id_sku == "80961" ~ 3.75,
      TRUE ~ unit_price
    )
  ) |> 
  filter(!is.na(unit_price))

# View the results
inventory_clean |> 
  filter(
    is.na(unit_price) | product_id_sku == "80961"
  ) |> 
  select(stock_date, product_id_sku, unit_price)
```

# Count calculations

The `category` column needs to be checked in case there are multiple variations of the same category. This could be an indication of misspelling or extra whitespace. Let's review in more detail.

```{r review-category-values}

# Find all the distinct values of category with their count of records
inventory_clean |> 
  count(category) |> 
  arrange(category)
```

By viewing the results in alphabetical order I can quickly spot spelling errors and extra whitespace in some of the category names.

```{r fix-category-spellings}

# Fix spelling errors in categories
inventory_clean <- 
  inventory_clean |> 
  mutate(category = str_trim(category)) |> 
  mutate(
    category = case_when(
      category == "Cookies Snacks & Candy" ~ "Cookies, Snacks & Candy",
      category == "Frozen Food" ~ "Frozen Foods",
      category == "Fruits & Vegtables" ~ "Fruits & Vegetables",
      TRUE ~ category
    )
  )

# View the results
inventory_clean |> 
  count(category) |> 
  arrange(category)
```

# Save the output

I have completed all of my preliminary QA steps. I can now save this clean dataframe to CSV so that I can use it for further profiling.

```{r save-output}
# Save the clean data
save(inventory_clean, file = "preliminary-qa.Rdata")
```

