---
title: "Data Profiling"
subtitle: "Univariate Profiling"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    theme: simplex
    toc: true
---

# Introduction

This is the second notebook in the series of data profiling steps. Make sure to run the first notebook before running this one. The steps taken in this notebook are:

-   Categorical distributions
-   Numerical distributions

```{r setup}
#| include: false

source("utils.R")
load("preliminary-qa.Rdata")
```

# Categorical distributions

Let's use `glimpse()` to see where the categorical variables are:

```{r find-categorical-variables}

glimpse(inventory_clean)
```

The categorical variables are `product_name`, `category` and `product_id_sku`. We can get an overview of these variables using `skimr`.

```{r skimr-categorical}
inventory_clean |> 
  select(where(is.character)) |> 
  skim()
```

Both `product_name` and `product_id_sku` have the same number of unique values as the number of rows in the dataframe, meaning that every record is unique. These are the primary keys of the dataframe, and therefore we do not need to visualise the distribution of their values.

## Category

Let's look at the frequency chart for `category`.

```{r frequency-chart-category}
# Frequency chart for category
inventory_clean |> 
  count(category, sort = TRUE) |> 
  mutate(prop = n / sum(n))
```

Here we can see there are 10 categories and the top value represents 17% of the data.

# Numerical distributions

Our numerical variables are `unit_price` and `units_sold`. We can also include the date value `stock_date` here. We can use `skimr` again to get some preliminary statistics.

```{r skimr-numerical}
inventory_clean |> 
  select(!where(is.character)) |> 
  skim()
```

We can view the distribution of each of these in a histogram and a density plot.

## Unit price

-   Max: `r round(max(inventory_clean$unit_price),2)`
-   Min: `r round(min(inventory_clean$unit_price),2)`
-   Mean: `r round(mean(inventory_clean$unit_price),2)`
-   Median: `r round(median(inventory_clean$unit_price),2)`
-   Variance: `r round(var(inventory_clean$unit_price),2)`
-   Standard deviation: `r round(sd(inventory_clean$unit_price),2)`
-   Skewness: `r round(skewness(inventory_clean$unit_price),2)`

```{r histogram-unit-price}

# Histogram for unit_price
inventory_clean |> 
  ggplot(
    aes(x = unit_price)
  ) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 1,
    fill = "#FF4F5C"
  ) +
  geom_density(
    colour = "darkslategray",
    linewidth = 1.5
  )
```

This variable seems to have an outlier at the higher end. Let's see if we can identify the record with the outlier.

```{r unit-price-outlier}

# Review the top 10 highest values for unit_price
inventory_clean |> 
  select(product_id_sku, product_name, category, unit_price) |> 
  slice_max(unit_price, n = 10)
```

It looks like there are 2 wine products that have a much higher unit price than all of the other products.

## Units sold

-   Max: `r round(max(inventory_clean$units_sold),2)`
-   Min: `r round(min(inventory_clean$units_sold),2)`
-   Mean: `r round(mean(inventory_clean$units_sold),2)`
-   Median: `r round(median(inventory_clean$units_sold),2)`
-   Variance: `r round(var(inventory_clean$units_sold),2)`
-   Standard deviation: `r round(sd(inventory_clean$units_sold),2)`
-   Skewness: `r round(skewness(inventory_clean$units_sold),2)`

```{r histogram-units-sold}

# Histogram for units sold
inventory_clean |> 
  ggplot(
    aes(x = units_sold)
  ) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 10,
    fill = "#FF4F5C"
  ) +
  geom_density(
    color = "darkslategray",
    linewidth = 1.5
  )
```

This distribution is also left-skewed with most values being less than 25.
